---
title: "Biostatistical Methods Homework 3"
output: word_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include = FALSE}
library(tidyverse)
library(knitr)
```

# Problem 1

##1

$$
\begin{aligned}
S^{2} &= \frac{1}{n-1}\sum_{i=1}^{n}\left(X_{i}^{2}+\bar{X}^{2}-2X_{i}\bar{X}\right)\\
&= \frac{1}{n-1}\left(\sum_{i=1}^{n}X_{i}^{2}+\sum_{i=1}^{n}\bar{X}^{2}-\sum_{i=1}^{n}2X_{i}\bar{X}\right)\\
&=\frac{1}{n-1}\left(\sum_{i=1}^{n}X_{i}^{2} + n\bar{X}^{2} - 2\bar{X}\sum_{i=1}^{n}X_{i}\right)\\
&= \frac{1}{n-1}\left(\sum_{i=1}^{n}X_{i}^{2} + n\bar{X}^{2} - 2\bar{X}n\bar{X}\right)\\
&= \frac{1}{n-1}\left(\sum_{i=1}^{n}X_{i}^{2} + n\bar{X}^{2} - 2n\bar{X}^{2}\right)\\
&= \frac{1}{n-1}\left(\sum_{i=1}^{n}X_{i}^{2} - n\bar{X}^{2} \right)\\
E(S^{2}) &= \frac{1}{n-1}\left [ E(\sum_{i=1}^{n}X_{i}^{2} )- E(n\bar{X}^{2}) \right ]\\
&= \frac{1}{n-1}\left [ nE(X_{i}^{2} )- nE(\bar{X}^{2}) \right ]\\
&= \frac{n}{n-1}\left [ E(X_{i}^{2} )- E(\bar{X}^{2}) \right ]\\
Var(\bar{X}) &= E\bar{X}^{2}-(E\bar{X})^{2}\\
E\bar{X}^{2} &= Var(\bar{X}) + (E\bar{X})^{2} \\
&= Var\left [ \frac{\sum_{i=1}^{n}X_{i}}{n} \right ]+\mu^{2}\\
\end{aligned}
$$
$$
\begin{aligned}
&=\frac{1}{n^{2}}Var\left ( \sum_{i=1}^{n}X_{i} \right )+\mu^{2}\\
&=\frac{1}{n^{2}}\sum_{i=1}^{n}Var(X_{i})+\mu^{2}\\
&=\frac{1}{n^{2}}n\sigma^{2}+\mu^{2}\\
&=\frac{1}{n}\sigma^{2}+\mu^{2}\\
Var(X_{i})&=EX_{i}^{2}-(EX_{i})^{2}\\
EX_{i}^{2}&=\sigma^{2}+\mu^{2}\\
E(S^{2})&=\frac{n}{n-1}\left[\sigma^{2}+\mu^{2}-(\frac{1}{n}\sigma^{2}+\mu^{2})\right]\\
&=\frac{n}{n-1}\left[\sigma^{2}-\frac{1}{n}\sigma^{2}\right]\\
&=\frac{n}{n-1}\left[\frac{n-1}{n}\sigma^{2}\right]\\
&=\sigma^{2}\\
\end{aligned}
$$



## 2
$$
\begin{aligned}
(y_{ij}-\bar{y})^{2}&=(y_{ij}-\bar{y_{i}}+\bar{y_{i}}-\bar{y})^{2}\\
&=(y_{ij}-\bar{y_{i}})^{2}+(\bar{y_{i}}-\bar{y})^{2}-2(y_{ij}-\bar{y_{i}})(\bar{y_{i}}-\bar{y})\\
\sum_{i=1}^{k}\sum_{j=1}^{n_{j}}(y_{ij}-\bar{y})^{2}&=\sum_{i=1}^{k}\sum_{j=1}^{n_{j}}(y_{ij}-\bar{y_{i}})^{2}+\sum_{i=1}^{k}\sum_{j=1}^{n_{j}}(\bar{y_{i}}-\bar{y})^{2}-\sum_{i=1}^{k}\sum_{j=1}^{n_{j}}2(y_{ij}-\bar{y_{i}})(\bar{y_{i}}-\bar{y})\\
\sum_{i=1}^{k}\sum_{j=1}^{n_{j}}(y_{ij}-\bar{y_{i}})(\bar{y_{i}}-\bar{y})&=\sum_{i=1}^{k}\sum_{j=1}^{n_{j}}(y_{ij}-\bar{y_{i}})\bar{y_{i}}-\sum_{i=1}^{k}\sum_{j=1}^{n_{j}}(y_{ij}-\bar{y_{i}})\bar{y}\\
&=0+0=0\\
\sum_{i=1}^{k}\sum_{j=1}^{n_{j}}(y_{ij}-\bar{y})^{2}&=\sum_{i=1}^{k}\sum_{j=1}^{n_{j}}(y_{ij}-\bar{y_{i}})^{2}+\sum_{i=1}^{k}\sum_{j=1}^{n_{j}}(\bar{y_{i}}-\bar{y})^{2}
\end{aligned}
$$


# Problem 2

Cigarette smoking continues to be a public health problem with major consequences on heart and lung diseases. Less is actually known about the consequences of quitting smoking. A recent study selected a group of 10 women working at a small medical practice, ages 50-64, that had smoked at least 1 pack/day and quit for at least 6 years (data “HeavySmoke.csv”).

## 1. The first question is to assess if their body mass index (BMI) has changed 6 years after quitting smoking. Perform an appropriate hypothesis test and interpret your findings. (5p)

```{r}
smoke_data = read_csv("./HeavySmoke.csv") %>%
  janitor::clean_names() %>%
  mutate(diff = bmi_base - bmi_6yrs)
```

```{r}
diff_mean = mean(smoke_data$diff)
diff_sd = sd(smoke_data$diff)
n = 10
t = (diff_mean - 0)/(diff_sd/sqrt(n))

```

```{r}
qt(0.975, n-1)
```

```{r}
t.test(smoke_data$bmi_base, smoke_data$bmi_6yrs, paired = TRUE)
```

$$
H_{0} : \mu_{1} - \mu_{2} = 0
$$

$$
H_{1} : \mu_{1} - \mu_{2} > 0
$$

$$
\bar{d}= \sum_{i=1}^{n}\frac{d_{i}}{n}=-3.36
$$

$$
s_{d}=\sqrt{\frac{\sum_{i=1}^{n}(d_{i}-\bar{d})^{2}}{n-1}}=2.4627
$$

$$
n = 10
$$

$$
t=\frac{\bar{d}-0}{s_{d}/\sqrt{n}}=-4.3145
$$

$$
t_{n-1, 1-\alpha /2}=2.262157
$$

$$
\left | t \right |=4.3145
$$

$$
For\ \left | t \right |> t_{n-1, 1-\alpha /2},\ reject\ H_{0}
$$
Intepretation:
We use paired t-test to test whether those 10 women's BMI has changed over 6 years after quitting smoking. According to the solutions listed above, we should reject the null, which means their BMI has changed significantly over 6 years.

## 2. The investigators suspected an overall change in weight over the years, so they decided to enroll a control group of 50-64 years of age that never smoked (data NeverSmoke.csv). Perform an appropriate test to compare the BMI changes between women that quit smoking and women who never smoked. Interpret the findings. (5p)

```{r}
nonsmoke_data = read_csv("./NeverSmoke.csv") %>%
  janitor::clean_names() %>%
  mutate(diff = bmi_base - bmi_6yrs)
```

```{r}
n1=10
n2=10
qf(0.975, n1-1, n2-1)
```

```{r}
#test equality for variances
var.test(smoke_data$diff, nonsmoke_data$diff, alternative = "two.sided")
```

$$
H_{0}: \sigma_{1} ^{2} = \sigma_{2} ^{2}
$$

$$
H_{0}: \sigma_{1} ^{2} \neq  \sigma_{2} ^{2}
$$

$$
F= \frac{s_{1}^{2}}{s_{2}^{2}}\sim F_{n_{1}-1,n_{2}-1}=1.162722 
$$

$$
F_{n_{1}-1,n_{2}-1} = 4.025994
$$

$$
For\ F< F_{n_{1}-1,n_{2}-1},\ fail\ to\ reject\ H_{0},\ \sigma_{1} ^{2} = \sigma_{2} ^{2}
$$

```{r}
qt(0.975, n1+n2-2)
```


```{r}
t.test(nonsmoke_data$diff, smoke_data$diff, var.equal = TRUE, paired = FALSE)
```

$$
H_{0} : \mu_{1} = \mu_{2}
$$

$$
H_{1} : \mu_{1} \neq \mu_{2}
$$

$$
s^{2} = \frac{(n_{1}-1)s_{1}^{2}+(n_{2}-1)s_{2}^{2}}{n_{1}+n_{2}-2}= 5.6405
$$

$$
t = \frac{\bar{X_{1}}-\bar{X_{2}}}{s\sqrt{(\frac{1}{n_{1}}+\frac{1}{n_{2}})}}=1.7041
$$

$$
t_{n_{1}+n_{2}-2, 1-\alpha /2}=2.100922
$$

$$
For\ \left | t \right |<t_{n_{1}+n_{2}-2, 1-\alpha /2},\ fail\ to\ reject\ H_{0},\ \mu_{1} = \mu_{2}
$$

Intepretation:
First, we use F-test to test the equality of variances. The result shows that the variances of two groups are equal. Then we use t-test to test the equality of mean. The result shows that the means of BMI changes of two groups are the same. So there is NOT a significant BMI changes between women who quit smoking and women who never smoked.

## 3. Show the corresponding 95% CI associated with part 2. Interpret it in the context of the problem.

$$
(\bar{X_{1}}-\bar{X_{2}}) - t_{n_{1}+n_{2}-2, 1-\alpha /2}s\sqrt{1/n_{1}+1/n_{2}}\leq \mu\leq (\bar{X_{1}}-\bar{X_{2}}) + t_{n_{1}+n_{2}-2, 1-\alpha /2}s\sqrt{1/n_{1}+1/n_{2}}
$$


```{r}
t = qt(0.975, 18)
s = sqrt(25.15739)
CIL = 28.86 - 30.41 - (t * s * sqrt(2/10))
CIR = 28.86 - 30.41 + (t * s * sqrt(2/10))
```

$$
(\bar{X_{1}}-\bar{X_{2}}) - t_{n_{1}+n_{2}-2, 1-\alpha /2}s\sqrt{1/n_{1}+1/n_{2}} = 
-6.262569
$$
$$
(\bar{X_{1}}-\bar{X_{2}}) + t_{n_{1}+n_{2}-2, 1-\alpha /2}s\sqrt{1/n_{1}+1/n_{2}} = 3.162569
$$
$$
-6.262569 \leq \mu \leq 3.162569
$$
Intepretation:
The 95% CI for these two samples are (-6.262569, 3.162569). This CI means that we are 95% confidence that the true population mean difference between women that quit smoking and women who never smoked lies between the lower and upper limits of the interval.

## 4. Suppose the researchers want to launch into a larger study to prove that a difference does exist between the two groups with respect to BMI changes.

### a. How would you design the new study? Comment on elements of study design such as randomization, possible causes of bias that should be avoided, etc. (5p)

For this new study, I would choose 50 women who never smoked and 50 women who quit smoking to conduct a cohort study. Make sure each individual in two groups are comparable. The possible bias in this study  should be avoided is that 1) we should have sufficient sample size. Greater sample size can better represent the population. If the sample size is too small, the result might be inaccurate; 2) make sure there is no missing values. 

### b. Calculate the sample size for the new study. Assuming a two-sided test, create a table showing sample size estimates for 80% vs 90% power, 2.5% vs 5% significance level, using the following information: the true mean increase for smokers is 3.0 kg/m2, with a standard deviation of 2.0 kg/m2; for never-smokers the true mean increase is 1.7 kg/m2, with a standard deviation of 1.5 kg/m2. (R only is allowed for calculations). (5p)

```{r}
sample_size <- function(power, alpha, x1, x2, sigma1, sigma2) {

  z_power <- qnorm(power)
  z_alpha <- qnorm(1-alpha/2)
  z_combine_square <- (z_power + z_alpha)^2
  diff <- x1-x2
  diff_square <- diff ^ 2
  sigma <- sigma1^2 + sigma2^2
  
  return(z_combine_square * sigma / diff_square)
  
 }
```

```{r}
sample_size(0.9, 0.05, 3, 1.7, 2, 1.5)
sample_size(0.8, 0.05, 3, 1.7, 2, 1.5)
sample_size(0.9, 0.025, 3, 1.7, 2, 1.5)
sample_size(0.8, 0.025, 3, 1.7, 2, 1.5)
```

$$
\Delta = 3-1.7=1.3
$$

$$
n = \frac{(z_{1-\beta} + z_{1-\alpha /2})^{2}(\sigma_{1}^{2}+\sigma_{2}^{2})}{\Delta ^{2}}
$$


Power       |    0.8     |     0.9    |
------------|------------|------------|
$\alpha$    |                         |
      0.025 |     35     |     46     |
       0.05 |     30     |     39     |



# Problem 3

A rehabilitation center is interested in examining the relationship between physical status before therapy and the time (days) required in physical therapy until successful rehabilitation. Records from patients 18-30 years old were collected and provided to you for statistical analysis (data “Knee.csv”).

Assuming that data are normally distributed, answer the questions below:


## 1. Generate descriptive statistics for each group and comment on the differences observed (R only). (4p)

```{r}
knee_data = read_csv("./Knee.csv") %>%
  janitor::clean_names()
```

```{r}
summary(knee_data$below)
sd(knee_data$below, na.rm = T)
summary(knee_data$average)
sd(knee_data$average, na.rm = T)
summary(knee_data$above)
sd(knee_data$above, na.rm = T)
```


Group       |    mean    |     sd     |   3rd Qu.  |
------------|------------|------------|------------|
      Below |     38     |  5.477226  |     42     |
    Average |     33     |  3.91578   |     35     |
      Above |    23.57   |  4.197505  |    24.5    |

As we can see above, the average days required in physical therapy until successful rehabilitation is largest in Below group and smallest in Above group. The number of 3rd Qu. is largest in the Below group and smallest in the Above group. These two findings are conform with our common sense. If the physical status before therapy is relatively better, the days required should be relatively shorter. 

## 2. Using a type I error of 0.01, obtain the ANOVA table. State the hypotheses, decision rule and conclusion (R only). (5p)

```{r}
below <- knee_data$below
average <- knee_data$average
above <- knee_data$above

knee_reshape <- c(below, average, above)
ind<-c(rep(3,length(below)),rep(2,length(average)),rep(1,length(above)))
new_data_knee <- as.data.frame(cbind(knee_reshape,ind))

res<-lm(knee_reshape~factor(ind), data=new_data_knee)
anova(res)
```

$$
H_{0}:\ \mu_{1} = \mu_{2} = \mu_{3} =\ ...\ =\mu_{k}
$$
$$
H_{1}:\ at\ least\ two\ means\ are\ not\ equal
$$
$$
F = \frac{Between\ SS/(k-1)}{With\ SS/(n-k)}\sim F_{k-1,n-k}\ distribution\ under\ H_{0}
$$
$$
F = 19.28
$$

```{r}
qf(1-0.01, 2, 22)
```


$$
F_{k-1,n-k,1-\alpha}=5.719022
$$
$$
F > F_{k-1,n-k,1-\alpha},\ reject\ H_{0}
$$
## 3. Based on your response in part 3, perform pairwise comparisons with the appropriate adjustments (Bonferroni, Tukey, and Dunnett – ‘below average’ as reference). Report your findings and comment on the differences/similarities between these three methods (R only). (5p)

### Bonferroni
```{r}
pairwise.t.test(new_data_knee$knee_reshape, new_data_knee$ind, p.adj='bonferroni')
qt( 1-((0.01/3)/2), 22 )
```
As we can see, according to the bonferroni adjustment, there are no mean differences between each group, which means:
$$
\mu_{1}=\mu_{2}=\mu_{3}
$$

### Tukey
```{r}
res1<-aov(knee_reshape~factor(ind), data=new_data_knee)
summary(res1)

TukeyHSD(res1, conf.level = 0.99)
```
According to the Tukey method, we can see that the mean between below and above and the mean between average and above are different. Tukey method is less conservative than Bonferroni.

### Dunnett
```{r}
library(DescTools)
```


```{r}
x <- c(29,42,38,40,43,40,30,42)
y <- c(30,35,39,28,31,31,29,35,39,33)
z <- c(26,32,21,20,23,22,21)
dunn_knee <- c(x,y,z)
g <- factor(rep(1:3, c(8, 10, 7)),
            labels = c("below",
                       "average",
                       "above"))
DunnettTest(dunn_knee, g, control = "above", conf.level = 0.99)
```
According to the Dunnett method, we can see that the mean between below and above group, and the mean between average and above group are different. This conclusion is consistent with the result using Tukey's method.


##4. Write a short paragraph summarizing your results as if you were presenting to the rehabilitation center director.(1p)

According to the analysis, we can conclude that for person who is in better physical status, the average days required in physical therapy is relatively less. However, the average days differences between below group and average group is not significant. We can conclude that the average days taken by patients in above group until rehabilitation is much less than patients in average group's and below group's. In a more conservative way, we might also says there is actually no significant differences between the average days taken by patients in each group.


# Problem 4

For this problem you will use the built-in R data called “UCBAdmissions” (library ‘datasets’), an example of sex bias in admission practices. You are interested in comparing the proportions of women vs men admitted at Berkeley (over all departments).

## 1. Provide point estimates and 95% CIs for the overall proportions of men and women admitted at Berkeley. Briefly comment on the values. (5p)


```{r}
library(datasets)
ucb_ad = as.data.frame(UCBAdmissions) %>%
  janitor::clean_names()
```

```{r}
ucb_women = ucb_ad %>%
  filter(gender == "Female")
ucb_men = ucb_ad %>%
  filter(gender == "Male")

ucb_admitted_women = ucb_women %>%
  filter(admit == "Admitted")
ucb_admitted_men = ucb_men %>%
  filter(admit == "Admitted")

X_men = sum(ucb_admitted_men$freq)
X_women = sum(ucb_admitted_women$freq)

n_men = sum(ucb_men$freq)
n_women = sum(ucb_women$freq)

p_hat_men = X_men/n_men
p_hat_women = X_women/n_women
```

```{r}
# CI for men
CIL_men = p_hat_men - (qnorm(0.975) * sqrt(p_hat_men * (1-p_hat_men)/n_men))
CIR_men = p_hat_men + (qnorm(0.975) * sqrt(p_hat_men * (1-p_hat_men)/n_men))

#CI for women
CIL_women = p_hat_women - (qnorm(0.975) * sqrt(p_hat_women * (1-p_hat_women)/n_women))
CIR_women = p_hat_women + (qnorm(0.975) * sqrt(p_hat_women * (1-p_hat_women)/n_women))
```

The calculation of CI for a population proportion:
$$
\hat{p}=\frac{\sum_{i=1}^{n}X_{i}}{n}=\frac{X}{n}
$$

$$
E(\hat{p}) = p
$$

$$
Var(\hat{p}) = \frac{p(1-p)}{n}
$$
$$
\hat{p} = \bar{X} \sim N(p,\frac{p(1-p)}{n})
$$
$$
(\hat{p} - z_{1-\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}},\hat{p} + z_{1-\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}})
$$

For men:
$$
\hat{p}=\frac{1198}{2691}=0.4451877
$$
$$
\hat{p} - z_{1-\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}=0.4264102
$$
$$
\hat{p} + z_{1-\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}=0.4639651
$$
So for men the CI is:
$$
(0.4264102,0.4639651)
$$

For women:

$$
\hat{p}=\frac{557}{1835}=0.3035422
$$

$$
\hat{p} - z_{1-\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}=0.2825051
$$
$$
\hat{p} + z_{1-\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}=0.3245794
$$
So for women the CI is:
$$
(0.2825051,0.3245794)
$$
Comment:
We are 95% confidence that the true population proportion of men admitted at Berkeley lies between 0.4264102 and 0.4639651, and the true population proportions of women admitted at Berkeley lies between 0.2825051 and 0.3245794. Simply judging from number, it seems that the CI of women is lower than CI of men. But we don't know for sure until we do the hypothesis test.


## 2. Perform a hypothesis test to assess if the two proportions in 1) are significantly different. Report the results including the test statistic and p-value and an overall conclusion of your findings. This part should contain both ‘hand’ and R calculations. For the latter, feel free to use built-in functions or to create your own. (5p)

```{r}
two.proptest_norm <- function(x1, x2, n1, n2, p=NULL, conf.level=0.95, alternative="less") {

# phat1, phat2 are observed proportions of each group
# n1, n2 are sample sizes of each group
# x1, x2 are admitted number of each group
# phat is the weighted average of the two sample proportions
# p.value is the hypothesis value
  
  z.stat <- NULL
  cint <- NULL
  p.val <- NULL
  phat1 <- x1/n1
  phat2 <- x2/n2
  qhat1 <- 1 - phat1
  qhat2 <- 1 - phat2
  phat <- (n1*phat1 + n2*phat2)/(n1+n2)
  qhat <- 1 - phat
  
 
  if(length(p) > 0) { 
    SE.phat <- sqrt(phat*qhat*((1/n1)+(1/n2))) 
    z.stat <- (phat1 - phat2)/SE.phat
    if(z.stat>0) {
      p.val <- pnorm(z.stat, lower.tail = FALSE)}
    if(z.stat<0) {
      p.val <- pnorm(z.stat, lower.tail = TRUE)
    }
      if(alternative=="two.sided") {
        p.val <- p.val * 2}
    
      if(alternative=="greater") {
        p.val <- 1 - p.val
      }
    } else {
    # Construct a confidence interval   
      SE.phat <- sqrt((phat1*qhat1/n1) + (phat2*qhat2/n2))
    }
    cint <- phat1-phat2 + c(-1*((qnorm(((1 - conf.level)/2) + conf.level))*SE.phat),
                        ((qnorm(((1 - conf.level)/2) + conf.level))*SE.phat))
  
  return(list(estimate=phat1-phat2, z.stat=z.stat, p.val=p.val, cint=cint))
}
```

```{r}
two.proptest_norm(x1=1198, x2=557, n1=2691, n2=1835, p=0, conf.level=0.95, alternative = "two.sided")
```

According to the result of two-sample binomial test for proportions, the test statistic is 9.602358, p-value is 7.8136e-22, confidence interval is (0.1127338, 0.1705571). Because the null hypothesis states that there is no differences between the proportions of each group, so the difference between two proportions should be 0. According to the confidence interval, we can see that 0 is not included. In this case, we should reject the null. We can double check this conclusion through p-value. The p-value is 7.8136e-22 which is way less than 0.05. In conclusion, the proportion of male admitted to UCB and the proportion of female admitted to UCB are significantly different. 

